Standardization of ceramic shape: A case studay from the Iron Age pottery from northeastern Taiwan

### 2020-08-02 analysis/paper/paper.Rmd
added morphological disparity test and updated the discussion section

```{r, first-two-PCs-byplot-and-tps-plot, echo=FALSE, fig.cap='Left: Significant differences in average vessel shapes across phases using thin plate splines (TPS) with outline deformations required to pass from an extreme of one morphospace to another. Right: Pottery shape distribution by each occupation phase according to the first two PCs.'}
knitr::include_graphics(here("analysis/figures/iso_pca_composit_crop.png"))
```

@@

To compare pottery shape standardization across the three phases we investigate the distributions of the first three PC scores, taking the PC scores as proxy variables for vessel shape (Figure \@ref(fig:PCs-distribution)). The first PC, capturing vessel height and roundness, shows higher variation in the pre-European period compared to the Chinese period. That indicates that shape standardization was higher in the Chinese period compared to the pre-European period. The second PC also presents a similar pattern of higher shape standardization in the Chinese period compared to the other two phases. The CVs calculated of the three PC also support a general trend toward a more standardized shape over time, especially for first PC.

To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we used a modified signed-likelihood ratio test to assess the equality of CVs [@Krishnamoorthy2014; @Marwick2019]. P-values for this significance test of CVs for PC1 and PC2 show significant differences in the standardization of vessel shapes across periods, especially between Chinese contact with either pre-European or post-European (Figure \@ref(fig:PCs-distribution), Table \@ref(tab:CV-test-PCs-table)). A significant difference was also detected between the pre-European and post-European periods.

@@

Previous investigations at Kiwulan suggested an unequal distribution of prestige goods, trade ornaments in particular, following the appearance of Europeans [@Cheng2008; @Wang2011], which hinted at the emergence of social inequality within the Indigenous community. Ceramic vessel shape standardization was examined to measure craft specialization as a proxy for social differentiation [@Costin2001; @Junker1999]. The result of our MANOVA demonstrates a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods. The average shape presents as a round body with a wide rim and neck before European contact, which shifts to a more oval-shaped body with narrower rim and neck after the European presence. Shape changes are more pronounced during the Chinese contact period. In general, vessels become shorter in height over time, leading to an oval-shaped body. This can be also supported by an increase in body diameter of vessels. Also, the correlation between shape and body diameter suggests that size significantly varies with vessel shape. Shorter vessels with a narrower opening tended to have a larger body diameter, indicative of a change in overall vessel form.

Our CV tests indicate that there are significant differences in shape standardization between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods. Pottery shape became more homogeneous and standardized after contact with Europeans and even more so after contact with the Chinese. In addition, we found a larger and more standardized body diameter of vessels in the Chinese period. We notice the possibility of more human errors could be involved in hand-crafting when the size of artifacts increases [@Eerkens2001]. However, it is opposite for the ceramics in the Chinese period when we examining body diameter. This might support the intent to achieve a homogeneous form of vessels, but we still need to compare it with other metric variables to assess the variation in size. Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909].

The spatial analysis (Figure \@ref(fig:kernel-plot)) for pottery samples presents multiple clusters with high densities of pottery during European presence. Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed distribution after European presence. In contrast, the distribution of pottery is more similar to random distributions during the Chinese period. This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups. The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population [@Chen2007; @Hsieh2009]. We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time. Although the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time. This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization. We also note that a smaller sample size in the Chinese period may have effects on a more standardized shape. The effect can be reduced using CV that offers a way to robustly compare variations across uneven sample sizes, regardless of a small sample size in the whole dataset [@Eerkens2001, pp.498]. Moreover, the MSLR test for equality of CVs can deal with different sample sizes as well. Also, the spatial pattern shows that Chinese pottery samples were collected from different units acorss research area.

>>>>

```{r, first-two-PCs-byplot-and-tps-plot, echo=FALSE, fig.cap='Left: Significant differences in average vessel shapes between the Chinese and the post-European period using thin plate splines (TPS) with outline deformations required to pass from an extreme of one morphospace to another. Right: Pottery shape distribution by each occupation phase according to the first two PCs.'}
knitr::include_graphics(here("analysis/figures/iso_pca_composit_crop.png"))
```

@@

To compare pottery shape standardization across the three phases we investigated the distributions of the first three PC scores, taking the PC scores as proxy variables for vessel shape (Figure \@ref(fig:PCs-distribution)). The CVs calculated of the three PC support a general trend toward a more standardized shape over time, especially the shape identified by PC1. The PC1, capturing vessel height and roundness, shows higher variation in the pre-European period and post-European period compared to the Chinese period. In other words, shape standardization was higher in the Chinese period. However, the PC2 presents a similar diversity in ceramics assemblages across three phases, while the PC3 demonstrates a slightly standardized shape in the Chinese period.

To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we used a modified signed-likelihood ratio test to assess the equality of CVs [@Krishnamoorthy2014; @Marwick2019]. P-values for this significance test of CVs for PC1 show significant differences in the standardization of vessel shapes across periods, between Chinese contact with either pre-European or post-European (Table \@ref(tab:CV-test-PCs-table)). This result statistically supports the observation of a higher standardized shape in Chinese period.

@@

Previous investigations at Kiwulan suggested an unequal distribution of prestige goods, trade ornaments in particular, following the appearance of Europeans [@Cheng2008; @Wang2011], which hinted at the emergence of social inequality within the Indigenous community. Ceramic vessel shape standardization was examined to measure craft specialization as a proxy for social differentiation [@Costin2001; @Junker1999]. The result of our MANOVA demonstrates a significant difference in shapes between the pre-European and the Chinese periods, and between the pre-European and Post-European periods. The average shape presents as a round body with a wide rim and neck before European contact, which shifts to a more oval-shaped body with narrower rim and neck after the European presence. Shape changes are more pronounced during the Chinese contact period. In general, vessels become shorter in height over time, leading to an oval-shaped body. This can be also supported by an increase in body diameter of vessels. Also, the correlation between shape and body diameter suggests that size significantly varies with vessel shape. Shorter vessels with a narrower opening tended to have a larger body diameter, indicative of a change in overall vessel form.

For the degree of shape standardization across periods, our CV tests on PC1 indicate that there are significant differences between Chinese contact and either pre-European or post-European periods. That means pottery shape became more homogeneous and standardized after contact with the Chinese. A test of morphological disparity we conducted as a cross validation that also supports this result. In addition, we found a larger and more standardized size, indicated by body diameter of vessels, in the Chinese period. We notice the possibility of more human errors could be involved in hand-crafting when the size of artifacts increases [@Eerkens2001]. However, it is opposite for the ceramics in the Chinese period using body diameter as a proxy variable of size. This might hint the intent to achieve a homogeneous form of vessels, but we still need to obtain more metric variables to assess the variation in size. Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909].

The spatial analysis (Figure \@ref(fig:kernel-plot)) for pottery samples presents multiple clusters with high densities of pottery during European presence. Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed distribution after European presence. In contrast, the distribution of pottery is more similar to random distributions during the Chinese period. This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups. The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population [@Chen2007; @Hsieh2009]. We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time. Although the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:kernel-plot) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time. This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization. We also note that a smaller sample size in the Chinese period may have effects on a more standardized shape. However, the effect can be reduced using CV, which offers a way to robustly compare variations across uneven sample sizes, regardless of a smaller sample size in the whole dataset [@Eerkens2001, pp.498]. Moreover, the MSLR test for equality of CVs can deal with different sample sizes as well.
